---
title: "`r params$report_title`"

output:
  bookdown::html_document2:
    css: report_styles.css
    number_sections: true
    toc: false
    fig_caption: true
    self_contained: true

params:
  area_of_interest: area_of_interest
  report_title: report_title
  sampling_coldspots: sampling_coldspots
---

```{r setup, message = FALSE, warning = FALSE, include = FALSE}

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 3,
  fig.height = 2,
  dpi = 300
)

library(tidyverse)
library(sf)
library(terra)
library(tmap)
library(gt)
library(patchwork)
library(cowplot)
library(bslib)
library(ranger)
library(pdp)
library(leaflet)
library(memoise)

source("biodiversity_trends_engine_functions.R")

tmap_mode("plot")

area_of_interest <- params$area_of_interest
sampling_coldspots <- params$sampling_coldspots

library(gt)

theme_report_table <- function(tbl){

  tbl |>

    opt_table_font(
      font = list(
        gt::google_font("Source Sans Pro")
      )
    ) |>

    tab_options(

      table.font.size = 13,

      heading.title.font.size = 18,

      heading.subtitle.font.size = 14,

      table.border.top.color = "#1F417D",

      table.border.bottom.color = "#1F417D",

      heading.border.bottom.color = "#4275D0",

      column_labels.border.bottom.color = "#4275D0",

      data_row.padding = px(5)

    ) |>

    opt_row_striping() |>

    tab_style(
      style = cell_text(
        weight = "bold",
        color = "#1F417D"
      ),
      locations = cells_column_labels()
    )

}
```

```{r extract-boundary-data, message=FALSE, warning=FALSE, include=FALSE}
# Filter boundary gbif data
area_of_interest$gbif_data_boundary <- area_of_interest$gbif_data[which(sf::st_intersects(area_of_interest$gbif_data, area_of_interest$boundary, sparse = FALSE)), ]
```

`r format(Sys.Date(), "%B %Y")`

# Key Insights

- Recording effort is unevenly distributed across habitats.
- Several habitat types remain underrepresented.
- Sampling coldspots reveal priority regions for future surveys.
- Patterns differ substantially among taxonomic groups.
- Habitat associations explain a significant component of trend variation.

# Underrepresented habitats

Underrepresented habitats refer to habitats which have been visited less frequently than would be expected given their prevalence across the conservation area of interest.

```{r underrepresented-habitats, include = TRUE, echo=FALSE, message=FALSE, warning=FALSE}
area_of_interest$underrepresented_habitats_df |>

  gt() |>

  fmt_percent(
    columns = where(is.numeric),
    decimals = 1
  ) |>

  theme_report_table()
```

# Sampling coldspots map

Sampling coldspots are locations across the conservation area of interest where undersampled habitats coincide with low overall rates of recording.

<div class="figure-panel">

```{r coldspots-map, include = TRUE, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
#### Where are the most underrepresented and undersampled habitats?
m <- leaflet::leaflet(options = leafletOptions(zoomDelta = 0.5, zoomSnap = 0, attributionControl = FALSE, worldCopyJump = FALSE)) %>% 
  leaflet::addProviderTiles(providers$Esri.WorldStreetMap) %>%
  leaflet::addRasterImage(
    sampling_coldspots, 
    colors = "#CD0600",
    opacity = 0.9
  ) %>% 
  leaflet::addPolygons(
    data = area_of_interest$boundary,
    color = grey(.2),
    fillOpacity = 0.1,
    fill = FALSE
  ) %>%
  leaflet::addLegend(
    position = "bottomleft", 
    colors = "#CD0600", 
    labels = "Sampling Coldspots \n (Underrepresented Habitats in Undersampled Areas"
  )

m
```
</div>

# Species associated with undersampled habitats

These are the 20 species which have most frequently been recorded in undersampled habitats but remain data deficient overall.

```{r coldspots-species, include=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
##### What species are associated with those habitats?
area_of_interest$underrepresented_habitats_species |>
  
  dplyr::select(-phylum, - genus) |>
  
  head(20) |>

  gt() |>

  cols_label(

    number_records = "Number of records",

    number_years_recorded = "Number of years recorded"

  ) |>

  theme_report_table()
```

# Under-represented/over-represented taxa

Representation refers to the degree of available information for a taxon given how frequently it occurs across the conservation area of interest. We estimate representation by assuming that the number of species ever recorded within a higher taxon is proportional to the likelihood of encountering them across the conservation area of interest. Under-represented and over-represented taxa are those which are recorded relatively less or more than expected given the number of species they include, respectively.

```{r undersampled-taxa, include=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
# Records representation
df <- area_of_interest$underrepresented_taxa %>% as.data.frame()

p1 <- ggplot(df, aes(group = taxon, x = proportion_species, y = proportion_records, fill = records_representation)) +
    ggplot2::geom_abline(intercept = 0, slope = 1) +
  geom_point(size = 2.5, colour = grey(.2), stroke = 0.2) +
     theme_linedraw() +
     theme(legend.position = "none",
           panel.grid = element_line(color = grey(.8)),
           axis.title = element_text(size = 9),
           axis.text = element_text(size = 8)
     ) +
  ggplot2::xlab("Proportion of all species in taxon") +
  ggplot2::ylab("Proportion of all records for taxon")

# Low sample size species representation
p2 <- ggplot(df, aes(group = taxon, x = proportion_low_sample_size_species, y = proportion_species, fill = low_sample_size_representation)) +
    ggplot2::geom_abline(intercept = 0, slope = 1) +
  geom_point(size = 2.5, colour = grey(.2), stroke = 0.2) +
     theme_linedraw() +
     theme(legend.position = "none",
           panel.grid = element_line(color = grey(.8)),
           axis.title = element_text(size = 9),
           axis.text = element_text(size = 8)
     ) +
  ggplot2::xlab("Proportion of all species in taxon") +
  ggplot2::ylab("Proportion of all undersampled species in taxon")

p <- subplot(p1, p2, nrows = 1, shareX = FALSE, titleX = TRUE, titleY = TRUE, margin = 0.05, which_layout = 1)

gg <- plotly::plotly_build(p) %>%
    plotly::config(displayModeBar = FALSE) %>%
    plotly::layout(# plot_bgcolor  = "rgba(0, 0, 0, 0)",
      paper_bgcolor = "rgba(0, 0, 0, 0)",
      font = list(family = "Helvetica", size = 14),
      xaxis = list(titlefont = list(size = 13),
                   tickfont = list(size = 13),
                   gridcolor = grey(.7)),
      yaxis = list(titlefont = list(size = 13),
                   tickfont = list(size = 13),
                   gridcolor = grey(.7))
    )
gg
```

# Largest trends in species' reporting rates

These are the (up to) 10 species showing the most positive and most negative trends in reporting rate anomaly. These species are being increasingly over- and under-reported compared to their historical rates of reporting at the same locations.

```{r biggest-movers, include=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
##### What species are associated with those habitats?
rbind(
        area_of_interest$biggest_movers_table %>%
          dplyr::filter(trend == "increasing") %>%
          dplyr::arrange(desc(reporting_rate_above_0_last_ten), desc(reporting_rate_trend_last_ten), desc(reporting_rate_trend_last_five)) %>%
          dplyr::slice_head(n = 10),
        area_of_interest$biggest_movers_table %>%
          dplyr::filter(trend == "decreasing") %>%
          dplyr::arrange(desc(reporting_rate_below_0_last_ten), reporting_rate_trend_last_ten, reporting_rate_trend_last_five) %>%
          dplyr::slice_head(n = 10)
      ) |>
            dplyr::select(-trend_icon) |>
  dplyr::distinct(species, .keep_all = TRUE) |>
          dplyr::mutate(
            reporting_rate_trend_last_five = sprintf("%.2f", reporting_rate_trend_last_five),
            reporting_rate_trend_last_ten = sprintf("%.2f", reporting_rate_trend_last_ten),
            trend = ifelse(trend == "increasing", "arrow-up", "arrow-down")
          ) |>
          dplyr::select(-reporting_rate_above_0_last_ten, -reporting_rate_below_0_last_ten) %>%
  
  gt() |>

  cols_label(
            species = "Species",
            trend = "Trend",
            number_records = "Total records",
            number_years_recorded = "Years recorded",
            reporting_rate_trend_last_five = "5-year trend",
            reporting_rate_trend_last_ten = "10-year trend"

  ) |>

  fmt_icon(
    columns = trend,
    fill_color = c("arrow-up" = "#67a9cf", "arrow-down" = "#ef8a62") 
  ) |>
  tab_style(
    style = cell_fill(color = "#67a9cf30"), 
    locations = cells_body(rows = trend == "arrow-up") 
  ) |>
  tab_style(
    style = cell_fill(color = "#ef8a6230"), 
    locations = cells_body(rows = trend == "arrow-down")
  ) |>
  theme_report_table()
```

# Average trends across taxa

These are the yearly trends in reporting rate anomaly averaged across all species in each major taxon for which a trend could be estimated. Values higher or lower than 0 represent instances of over-reporting or under-reporting across all species in the taxon in a given year.

```{r taxon-summed-trends, include=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.height=7, fig.width=12, out.width='100%'}
taxon_list <- area_of_interest$major_taxon_trends_summed[c("Aves", "Mammalia", "Squamata", "Amphibia", "Insecta")]
df <- purrr::map(1:length(taxon_list), function(t){
  dat <- taxon_list[[t]] %>% 
      dplyr::select(year, mean_metric) %>%  
      dplyr::mutate(taxon = names(taxon_list)[t])
  }
) %>% 
  bind_rows() %>% 
  as.data.frame()
p3 <- ggplot(df,
             aes(year, mean_metric, colour = taxon)) +
  geom_hline(yintercept = 0,
             colour = "black",
             linewidth = 1) +
  geom_line(linewidth = .7) +
  facet_wrap(
    vars(taxon),
    scales = "free_y",
    ncol = 3
  ) +
  theme_linedraw() +
  theme(
    legend.position = "none",
    panel.grid = element_line(color = grey(.8)),
    strip.text = element_text(size = 16),
    axis.text.y = element_text(size = 15),
    axis.text.x = element_text(size = 15, angle = 45, hjust = 1),
    axis.title = element_text(size = 15)
  ) +
  labs(
    x = NULL,
    y = "Reporting rate anomaly"
  )
p3

# gg <- plotly::plotly_build(p3) %>%
#   plotly::config(displayModeBar = FALSE) %>%
#   plotly::layout(# plot_bgcolor  = "rgba(0, 0, 0, 0)",
#     paper_bgcolor = "rgba(0, 0, 0, 0)",
#     font = list(family = "Helvetica", size = 14),
#     xaxis = list(titlefont = list(size = 13),
#                  tickfont = list(size = 13),
#                  gridcolor = grey(.7)),
#     yaxis = list(titlefont = list(size = 13),
#                  tickfont = list(size = 13),
#                  gridcolor = grey(.7))
#   )
# 
# gg
```

# Average trends across habitats

These are the yearly trends in reporting rate anomaly averaged across all species with an estimated trend associated with each habitat. Values higher or lower than 0 represent instances of over-reporting or under-reporting across species associated with that habitat in a given year.

```{r habitat-summed-trends, include=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.height=7, fig.width=12, out.width='100%'}
df <- purrr::map(1:length(area_of_interest$major_habitat_trends), function(t){
  dat <- area_of_interest$major_habitat_trends[[t]] %>% 
      dplyr::select(year, mean_metric) %>%  
      dplyr::mutate(taxon = names(area_of_interest$major_habitat_trends)[t])
  }
) %>% 
  bind_rows() %>% 
  as.data.frame()
p4 <- ggplot(df, aes(x=year, y = mean_metric, colour = taxon)) +
    geom_hline(yintercept = 0, colour = "black", size = 1) +
    # geom_hline(yintercept = 1.96, colour = grey(.5), size = .5) +
    # geom_hline(yintercept = -1.96, colour = grey(.5), size = .5) +
    geom_line(size = 0.7, alpha = 1) +
  facet_wrap(
    vars(taxon),
    scales = "free_y",
    ncol = 3
  ) +
  theme_linedraw() +
  theme(
    legend.position = "none",
    panel.grid = element_line(color = grey(.8)),
    strip.text = element_text(size = 16),
    axis.text.y = element_text(size = 15),
    axis.text.x = element_text(size = 15, angle = 45, hjust = 1),
    axis.title = element_text(size = 15)
  ) +
  labs(
    x = NULL,
    y = "Reporting rate anomaly"
  )

p4
# 
#   gg <- plotly::plotly_build(p4) %>%
#     plotly::config(displayModeBar = FALSE) %>%
#     plotly::layout(# plot_bgcolor  = "rgba(0, 0, 0, 0)",
#       paper_bgcolor = "rgba(0, 0, 0, 0)",
#       font = list(family = "Helvetica", size = 14),
#       xaxis = list(titlefont = list(size = 13),
#                    tickfont = list(size = 13),
#                    gridcolor = grey(.7)),
#       yaxis = list(titlefont = list(size = 13),
#                    tickfont = list(size = 13),
#                    gridcolor = grey(.7))
#     )
#   gg
```


# Predictors of trends: relative importance

These are the main taxonomic and habitat drivers of species' yearly trends resulting from a predictive machine learning model (random forest).

```{r predictor-importance, include=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.height=7, fig.width=12, out.width='100%'}
# Importance
imp <- sort(importance(area_of_interest$rf), decreasing = FALSE) %>% stack()

ggplot(imp, aes(x = ind, y = values)) +
  geom_col(fill = "#4275D0") +
  coord_flip() +
  theme_linedraw() +
  theme(
    legend.position = "none",
    panel.grid = element_line(color = grey(.8)),
    strip.text = element_text(size = 16),
    axis.text.y = element_text(size = 15),
    axis.text.x = element_text(size = 15, angle = 45, hjust = 1),
    axis.title = element_text(size = 15)
  ) +
  labs(
    y = "Variable Importance",
    x = ""
  )
```

# Predictors of trends: taxon partial effects

These are the strongest positive and negative relationships between taxonomic groups and yearly trend. Categories with values above or below 0 are associated with more positive or more negative yearly trends, respectively.

```{r taxon-partial-effects, include=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.height=7, fig.width=12, out.width='100%'}
get_partials_byTaxon <- function(taxonomic_scale = "phylum"){
  
  taxon_plot <- partial(
    area_of_interest$rf,
    pred.var = taxonomic_scale,
    train = area_of_interest$rf_data, 
    plot = TRUE, 
  )
  
  taxon_df <- data.frame(taxa = as.factor(taxon_plot$panel.args[[1]]$x),
                         trend = taxon_plot$panel.args[[1]]$y
  ) 
  
  freq <- table(area_of_interest$rf_data[[taxonomic_scale]])
  taxon_df_extremes <- taxon_df %>% 
    mutate(n = freq[taxa]) %>%
    filter(n >= 3) %>%
    dplyr::filter(
      trend < quantile(trend, .25) | trend > quantile(trend, .75)
    ) 
  taxon_df_extremes <- taxon_df_extremes %>% 
    dplyr::mutate(
      taxa = factor(taxa, levels = taxon_df_extremes %>% dplyr::arrange(desc(trend)) %>% dplyr::pull(taxa))
    )
  
  ggplot(taxon_df_extremes, aes(x = taxa, y = trend)) +
  geom_col(fill = "#4275D0") +
  coord_flip() +
  theme_linedraw() +
theme(
    legend.position = "none",
    panel.grid = element_line(color = grey(.8)),
    strip.text = element_text(size = 16),
    axis.text.y = element_text(size = 15),
    axis.text.x = element_text(size = 15, angle = 45, hjust = 1),
    axis.title = element_text(size = 15)
  ) +
  labs(
    y = "Trend",
    x = ""
  )
  
}
get_partials_byTaxon(taxonomic_scale = "class")
```

# Predictors of trends: habitat partial effects

These are the strongest positive and negative relationships between habitats and yearly trend. Categories with values above or below 0 are associated with more positive or more negative yearly trends, respectively.

```{r habitat-partial-effects, include=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.height=7, fig.width=12, out.width='100%'}
get_partials_byHabitat <- function(habitats = gsub(" |-", "_", area_of_interest$major_habitats)){
  
  habitat_partials <- purrr::map(1:length(habitats), function(hab){
    
    habitat_plot <- pdp::partial(
      area_of_interest$rf,
      pred.var = habitats[hab],
      train = area_of_interest$rf_data, 
      plot = TRUE, 
    )
    
    freq <- table(area_of_interest$rf_data[[habitats[hab]]])
    
    habitat_df <- data.frame(habitat = as.factor(habitat_plot$panel.args[[1]]$x),
                             trend = habitat_plot$panel.args[[1]]$y
    ) %>% 
      mutate(n = freq[habitat]) %>%
      dplyr::filter(habitat == 1) %>% 
      dplyr::mutate(habitat = gsub("_", " ", habitats[hab]))
    
   if (nrow(habitat_df) == 0) habitat_df <- NULL
    
    habitat_df
    
  }) %>% 
    bind_rows() %>% 
    purrr::compact()
  
    habitat_partials <- habitat_partials %>% 
    dplyr::mutate(
      habitat = factor(habitat, levels = habitat_partials %>% dplyr::arrange(desc(trend)) %>% dplyr::pull(habitat))
    )
  
  ggplot(habitat_partials, aes(x = habitat, y = trend)) +
  geom_col(fill = "#4275D0") +
  coord_flip() +
  theme_linedraw() +
    theme(
      legend.position = "none", 
      panel.grid = element_line(color = grey(.8)),
      strip.text = element_text(size = 16),
    axis.text.y = element_text(size = 15),
    axis.text.x = element_text(size = 15, angle = 45, hjust = 1),
    axis.title = element_text(size = 15)
    ) +
  labs(
    y = "Trend",
    x = ""
  )
  
}

get_partials_byHabitat()
```
